### 2021-07-13 src/paper/model/param_tables.tex
Nicely format tables.
Minor improvements.
Merge tables.
Shortened some parts of main text, added discussion of rapid tsts.

@@

\begin{table}
\centering
\caption{contacts_and_matching}
\label{tab:contacts_and_matching}
\input{../bld/param_tables/contacts_and_matching}
\end{table}

\begin{table}
\centering
\caption{policies}
\label{tab:policies}
\input{../bld/param_tables/policies}
\end{table}

\begin{table}
\centering
\caption{infection_probs}
\label{tab:infection_probs}
\input{../bld/param_tables/infection_probs}
\end{table}

\begin{table}
\centering
\caption{disease_model}
\label{tab:disease_model}
\input{../bld/param_tables/disease_model}
\end{table}

\begin{table}
\centering
\caption{pcr_testing_and_case_detection}
\label{tab:pcr_testing_and_case_detection}
\input{../bld/param_tables/pcr_testing_and_case_detection}
\end{table}

\begin{table}
\centering
\caption{rapid_testing}
\label{tab:rapid_testing}
\input{../bld/param_tables/rapid_testing}
\end{table}

\begin{table}
\centering
\caption{behavioral_response}
\label{tab:behavioral_response}
\input{../bld/param_tables/behavioral_response}
\end{table}

\begin{table}
\centering
\caption{vaccinations}
\label{tab:vaccinations}
\input{../bld/param_tables/vaccinations}
\end{table}

>>>>

\clearpage

\begin{landscape}

@@

\comment[id=HM]{Copied from main text. Do we describe the parameter for fraction of
vaccination-induced immunity/sterility  somewhere? I could not find it. Nothing special
about the 75\%...

Also, start out more positively. Something like ``We do not distinguish between first
and second doses for computational reasons. We thus use a weighted average ...''

75\% is lower than what is usually reported for after the second dose
of the Biontech/Pfizer vaccine, which is most commonly used in Germany. We choose it
because our model neither includes booster shots, nor does it allow vaccinated
individuals who became immune to transmit the disease\citep{Petter2021,
LevineTiefenbrun2021, Pritchard2021}. If anything, these assumptions would overstate the
effect of vaccines for our study period. This would be different if a large fraction of
vaccinated individuals had received a second dose already.

}

\FloatBarrier

\begin{table}[htb]
\centering
\caption{Contacts, Matching and Policies}
\label{tab:contacts_and_matching}
\input{../bld/param_tables/contacts_matching_policies}
\end{table}

\begin{table}[htb]
\centering
\caption{Infection Probabilities and Virus Variants}
\label{tab:infection_probs}
\input{../bld/param_tables/infection_probs}
\end{table}

\begin{table}[htb]
\centering
\caption{Disease and Vaccination Model}
\label{tab:disease_and_vacc_model}
\input{../bld/param_tables/disease_model}
\end{table}

\begin{table}[htb]
\centering
\caption{Rapid Testing}
\label{tab:rapid_testing}
\input{../bld/param_tables/rapid_testing}
\end{table}

\begin{table}[htb]
\centering
\caption{PCR Testing, Case Detection and Behavioral Response}
\label{tab:pcr_testing_and_case_detection}
\input{../bld/param_tables/pcr_testing_and_case_detection}
\end{table}

\FloatBarrier

\end{landscape}

\FloatBarrier

### 2021-07-13 src/paper/report.tex
Shortened some parts of main text, added discussion of rapid tsts.

At the core of our agent-based model are physical contacts between heterogeneous agents
(Figure~\ref{fig:model_contacts_infections}).\footnote{A detailed comparison with other
approaches  is relegated to Supplementary Material~\ref{sec:literature_review}. The model
most closely related to ours is described in \citet{Hinch2020}.} Each contact between an
infectious individual and somebody susceptible to the disease bears the risk of
transmitting the virus\comment[id=K]{infection probability per contact for contact type
$c$: $\beta_{c}$}. Contacts occur in up to four networks: Within the household, at work,
at school, or in other settings (leisure activities, grocery shopping, medical
appointments, etc.). Some contacts recur regularly, others occur at random. Empirical
applications can take the population and household structure from census data and the
network-specific frequencies of contacts\comment[id=K]{probability to have $n$ contacts
of contact type $c$: $\eta_{c,\:n}$} from diary data measuring contacts before the
pandemic \citep[e.g.][]{Mossong2008,Hoang2019}. Within each network, meeting frequencies
depend on age and geographical location (see Supplementary Material
Section~\ref{subsec:assortativity}).\comment[id=K]{Share of contacts of age group and
county of i to meet someone of j's age group and county: $\alpha_{work,\:age_i,
age_j,\:county_i,\:county_j}$}

@@

In the model, susceptibility to contracting the SARS-CoV-2 virus is dependent on
age\comment[id=K]{$\zeta_{a}$}. A possible infection progresses as shown in
Figure~\ref{fig:model_disease_progression}.\comment[id=HM]{New version of figure has
lost the timing. Need to explain d-subscript, maybe even better to add stuff like 1-3
days back in? Needs to be comprehensible without recurring to appendix} We differentiate
between an initial period of infection without being infectious or showing symptoms,
being infectious (presymptomatic or asymptomatic), showing symptoms, requiring intensive
care, and recovery or death \citep[similar to][]{Grimm2021}.\comment[id=HM]{Need to look
for non-economists -- this must be fairly standard?} The probabilities of transitioning
between these states depend on age; their duration is random and calibrated to medical
literature (for a detailed description see Section~\ref{sub:data_course_of_disease}).
Conditional on the type of contact, infectiousness is independent of age
\citep{Jones2021}.\comment[id=K]{would be $\beta_{c}$}

The model includes several other features, which are crucial to describe the evolution
of the pandemic in 2020-2021. New virus strains with different profiles regarding
infectiousness\comment[id=K]{$\sigma_v$} can be introduced. Agents may receive a
vaccination. With a probability of 75\% \citep{Hunter2021}, vaccinated agents become
immune and they do not transmit the virus \citep{Petter2021, LevineTiefenbrun2021,
Pritchard2021}.\footnote{75\% is lower than what is usually reported for after the
second dose of the Biontech/Pfizer vaccine, which is most commonly used in Germany. We
choose it because our model neither includes booster shots, nor does it allow vaccinated
individuals who became immune to transmit the disease\citep{Petter2021,
LevineTiefenbrun2021, Pritchard2021}. If anything, these assumptions would overstate the
effect of vaccines for our study period. This would be different if a large fraction of
vaccinated individuals had received a second dose already.} During the vaccine roll-out,
priority may depend on age and occupation; some individuals will refuse vaccination
offers.

We include two types of tests. Polymerase chain reaction (PCR) tests reveal whether an
individual is infected or not; there is no uncertainty to the result. PCR tests require
at least one day to be processed\comment[id=K]{$\gamma_{PCR,\:d}$ where $d=$ number of
days} and there are aggregate capacity constraints. In contrast, rapid antigen tests
yield immediate results; after a phase-in period, all tests that are demanded will be
performed. Specificity\comment[id=K]{$p_{negative|not\:infected}$} and
sensitivity\comment[id=K]{"$p_{positive|infected,\;i,\:t}$", where $i$ is the individual
and $t$ is the date to determine time until/since infectiousness} of these tests is set
according to data analyzed in \cite{Bruemmer2021, Smith2021}; sensitivity depends on the
timing of the test relative to the onset of infectiousness.
Figure~\ref{fig:model_rapid_tests} shows our model for rapid test demand. Schools may
require staff and students to be tested regularly.\comment[id=K]{"$\pi_{teacher,\:t}$"
etc. } Rapid tests may be offered by employers to on-site workers. Individuals may
demand tests for private reasons, which include having plans to meet other people,
showing symptoms of CoViD-19, and because a household member tested positively for the
virus. We endow each agent with an individual compliance parameter. This parameter
determines whether she takes up rapid tests.\footnote{A positive test result (of either
kind), as well as symptoms leads most individuals to reduce their contacts; this is why
tests impact the actual contacts in Figure~\ref{fig:model_description}.}

Modelling a population of agents according to actual demographic characteristics means
that we can use a wide array of data to identify and estimate the model's many
parameters.\footnote{See section~\ref{sec:data_and_parameters} of the supplementary
materials for the precise sources.} Contact diaries yield pre-pandemic distributions of
contacts for different contact types and their assortativity by age group. Mobility data
is used to model the evolution of work contacts. School and daycare policies can be
incorporated directly from official directives. Administrative records on the number of
tests, vaccinations by age and region, and the prevalence of virus strains are generally
available. Surveys may ask about test offers, propensities to take them up, and past
tests. Other studies' estimates of the seasonality of infections can be incorporated
directly. The remaining parameters---most notably, these include infection probabilities
by contact network and the effects of some NPIs, see Supplementary
Material~\ref{subsec:estimated_params}---will be chosen numerically so that the model
matches features of the data \citep[see][for the general method, also described in
Supplementary Material \ref{subsec:estimated_params}]{McFadden1989}. In our application,
we keep the number of free parameters low in order to avoid overfitting. The data
features to be matched include official case numbers for each age group and region,
deaths, and the share of the B.1.1.7 strain.

The main issue with official case numbers is that they will contain only a fraction of
all infections. We thus model official cases as depicted in
Figure~\ref{fig:model_official_cases}\comment[id=HM]{Replace P2, P(compliance) by
symbols in figure}. We take aggregate estimates of the share of detected
cases\comment[id=K]{$\psi$} and use data on the share of PCR tests administered because
of CoViD-19 symptoms\comment[id=K]{$\chi_{symptom}$}. As the share of asymptomatic
individuals varies by age group, this gives us age-specific shares (see
Figure~\ref{fig:share_known_cases_by_age_group} for the share of detected cases by age
group over time in our model). Our estimates suggest that---in the absence of rapid
testing---the detection rate is 80\% higher on average for individuals above age 80
compared to school age children. Once rapid test become available, confirmation of a
positive result\comment[id=K]{prob of demanding confirmation is $\chi_{confirmation}$}
is another reason leading to positive PCR tests.

The model is applied to the second and third waves of the CoViD-19 pandemic in Germany,
covering the period mid-September 2020 to the end of May 2021.
Figure~\ref{fig:pandemic_drivers_model_fit} describes the evolution of the pandemic and
of its drivers. The black line in Figure~\ref{fig:aggregated_fit} shows officially
recorded cases; the black line in Figure~\ref{fig:stringency_infectious_contacts} the
Oxford Response Stringency Index \citep{Hale2020}, which tracks the tightness of
non-pharmaceutical interventions. The index is shown for illustration of the NPIs, we
never use it directly. For legibility reasons, we transform the index so that lower
values represent higher levels of restrictions. A value of zero means all measures
incorporated in the index are turned on. The value one represents the situation in
mid-September, with restrictions on gatherings and public events, masking requirements,
but open schools and workplaces. In the seven weeks between mid September and early
November, cases increased by a factor of ten. Restrictions were somewhat tightened in
mid-October and again in early November. New infections remained constant throughout
November before rising again in December, prompting the most stringent lockdown to this
date. Schools and daycare centers were closed, so were customer-facing businesses except
for grocery and drug stores. From the peak of the second wave just before Christmas until
the trough in mid-February, newly detected cases decreased by almost three quarters. The
third wave in the spring of 2021 is associated with the B.1.1.7 strain, which became
dominant in March (Figure~\ref{fig:share_b117}) and stayed dominant for our entire
simulation period.\footnote{B.1.617.2 (Delta) reached Germany in April but still
accounted for less than 5\% of cases at the end of our simulation period.}
In early March, some NPIs were being relaxed; e.g., hairdressers and home improvement
stores were allowed to open again to the public. There were many changes in details of
regulations afterwards, but they did not change the overall stringency index.

@@

By March 2021, the set of policy instruments had become much more diverse. Around the
turn of the year, the first people were vaccinated with a focus on older age groups and
medical staff (Figure~\ref{fig:antigen_tests_vaccinations}). Until the end of May, 43\%
had received at least one dose of a vaccine. In late 2020, rapid tests started to
replace regular PCR tests for staff in many medical and nursing facilities. These had to
be administered by medical doctors or in pharmacies. At-home tests approved by
authorities became available in mid-March, rapid test centers were opened, and one test
per person and week was made available free of charge. In several states, customers were
only allowed to enter certain stores with a recent negative rapid test result. These
developments are characteristic of many countries: The initial focus on NPIs to slow the
spread of the disease has been accompanied by vaccines and a growing acceptance and use
of rapid tests. At broadly similar points in time, novel strains of the virus have
started to pose additional challenges.

We draw simulated samples of agents from the population structure in September 2020 and
use the model to predict recorded infection rates until the end of May 2021. See
Supplementary Material sections~\ref{subsec:synthetic_population} and
\ref{sub:initial_conditions} for details. The blue line in
Figure~\ref{fig:aggregated_fit} shows that our model's predictions are very close to
officially recorded cases in the aggregate. This is also true for infections by age and
geographical region, which are shown in the supplementary materials
(Figures~\ref{fig:age_group_fit} and \ref{fig:state_fit}, respectively).

The effects of various mechanisms can be disentangled due to the distinct temporal
variation in the drivers of the pandemic. Next to the stringency index, the three lines
in Figure~\ref{fig:stringency_infectious_contacts} summarize how contact reductions,
increased hygiene regulations, and seasonality evolved since early September for each of
the three broad contact networks. For example, a value of 0.75 for the work multiplier
means that if the environment was the same as in September (levels of infection rates,
no rapid tests or vaccinations, only the wildtype virus present), infections at the
workplace would be reduced by 25\%. The lines show the product of the effect of contact
reductions, increased hygiene regulations, and seasonality. Two aspects are particularly
interesting. First, all lines broadly follow the stringency index and they would do so
even more if we left out seasonality and school vacations (roughly the last two weeks of
October, two weeks each around Christmas and Easter, and some days in late May). Second,
the most stringent regulations are associated with the period of strong decreases in new
infections between late December 2020 and mid-February 2021. The reversal of the trend
is associated with the spread of the B.1.1.7 variant. The steep drop in recorded cases
during May 2021 is associated with at least weekly rapid tests for 42\% of the
population, a vaccination rate that rose from 28\% to 43\%, and seasonality lowering the
relative infectiousness of contacts, most strongly for other contacts.

In order to better understand the contributions of rapid tests, vaccinations, and
seasonality on the evolution of infections in 2021, Figure~\ref{fig:2021_scenarios_broad}
considers various scenarios. NPIs are always the same as in the baseline scenario.
Figure~\ref{fig:2021_scenarios_recorded} shows the model fit (the blue line, same as in
Figure~\ref{fig:aggregated_fit}), a scenario without any of the three factors (red line),
and three scenarios turning each of these factors on individually.
Figure~\ref{fig:2021_scenarios_newly_infected} does the same for total infections in the
model. Figure~\ref{fig:2021_scenarios_decomposition} employs Shapley values
\citep{Shapley2016} to decompose the difference in total infections between the scenario
without any of the three factors and our main specification.

@@

Until mid-March, there is no visible difference between the different scenarios.
Seasonality hardly changes, and only few vaccinations and rapid tests were administered.
Even thereafter, the effect of the vaccination campaign is surprisingly small at first
sight. Whether considering recorded or total infections with only one channel active,
the final level is always the highest in case of the vaccination campaign (orange
lines). The Shapley value decomposition shows that vaccinations contribute 16\% to the
cumulative difference between scenarios. Reasons for this are the slow start---it took
until March~24th until 10\% of the population had received their first vaccination, the
20\% mark was reached on April 19th---and the focus on older individuals. These groups
contribute less to the spread of the disease than others due to a lower number of
contacts. By the end of our study period, when first-dose vaccination rates reached
around 40\% of the population, the numbers of new cases would have started to decline.
It is important to note that the initial focus of the campaign was to prevent deaths and
severe disease. Indeed, the case fatality was rate considerably lower during the third
wave when compared to the second (4.4\% between October and February and 1.4\% between
March and the end of May).\comment[id=K]{This needs a source.}\comment[id=HM]{Didn't we
calculate that ourselves at some point?}

@@

A similar-sized effect---42\% in the decomposition---comes from rapid testing. Here, it
is crucial to differentiate between recorded cases and actual cases. Additional testing
means that additional infections will be recorded which would otherwise remain
undetected. Figure~\ref{fig:2021_scenarios_recorded} shows that this effect is large.
Until late April, recorded cases are higher in the scenario with rapid testing alone when
compared to the setting where none of the three mechanisms are turned on. The effect on
total cases, however, is visible immediately in
Figure~\ref{fig:2021_scenarios_newly_infected}. Despite the fact that only 10\% of the
population performed weekly rapid tests in March on average, new infections on April 1
would have been reduced by 53\% relative to the scenario without vaccinations, rapid
tests, or seasonality.

So why is rapid testing so effective? In order to shed more light on this question,
Figure~\ref{fig:2021_scenarios_decomposition_tests} decomposes the difference in the
scenario without rapid tests only and the main specification into the three channels for
rapid tests. Tests at schools have the smallest effect, which is largely explained by
schools not operating at full capacity during our period of study and the relatively
small number of students.\footnote{18\% of our population are in the education sector
(pupils, teachers, etc.); 46\% are workers outside the education sector.} Almost 40\%
come from tests at the workplace. Despite the fact that rapid tests for private reasons
are phased in quite late, they make up for more than half of the total effect. The reason
lies in the fact that a substantial share of these tests is driven by an elevated
probability to carry the virus, i.e., showing symptoms of CoViD-19 or following up on a
positive test of a household member. The latter is essentially a form of contact tracing,
which has been shown to be very effective \citep{Contreras2021,
Kretzschmar2020,Fetzer2020}.

@@

Our analysis has shown that during the transition to high levels of vaccination and
possibly thereafter, large-scale rapid testing can substitute for some NPIs. This comes
at a fraction of the cost. A week of the fairly strict lockdown in early 2021 is
estimated to have cost around 20~Euros per capita \citep{Wollmershauser2021}; retail
prices for rapid tests were below one Euro in early June 2021. Despite these large
effects, the results on testing likely understate the benefits. Disadvantaged groups are
less likely to be reached by testing campaigns relying on voluntary participation
\citep[e.g.][]{StillmanTonin2021}; at the same time, these groups have a higher risk to
contract CoViD-19 \citep{KochInstitut2021a}. Mandatory tests at school and at the
workplace will extend more into these groups. The same goes for individuals who exhibit
a low level of compliance with CoViD-19-related regulations. Compared to vaccinations,
rapid testing programmes allow a much quicker roll-out, making it arguably the most
effective tool to contain the pandemic in the short run.

>>>>

At the core of our agent-based model \citep[][we review more literature in Supplementary
    Material~\ref{sec:literature_review}]{Aleta2020,Hinch2020} are physical contacts
between heterogeneous agents (Figure~\ref{fig:model_contacts_infections}). Each contact between an infectious individual and somebody susceptible to
the disease bears the risk of transmitting the virus\comment[id=K]{infection probability
per contact for contact type $c$: $\beta_{c}$}. Contacts occur in up to four networks:
Within the household, at work, at school, or in other settings (leisure activities,
grocery shopping, medical appointments, etc.). Some contacts recur regularly, others
occur at random. Empirical applications can take the population and household structure
from census data and the network-specific frequencies of
contacts\comment[id=K]{probability to have $n$ contacts of contact type $c$:
$\eta_{c,\:n}$} from diary data measuring contacts before the pandemic
\citep[e.g.][]{Mossong2008,Hoang2019}. Within each network, meeting frequencies depend
on age and geographical location (see Supplementary
Material~\ref{subsec:assortativity}). \comment[id=K]{Share of contacts of age group and
county of i to meet someone of j's age group and county: $\alpha_{work,\:age_i,
        age_j,\:county_i,\:county_j}$}

@@

In the model, susceptibility to contracting the SARS-CoV-2 virus is dependent on
age\comment[id=K]{$\zeta_{a}$}. A possible infection progresses as shown in
Figure~\ref{fig:model_disease_progression}.\comment[id=HM]{New version of figure has
lost the timing. Need to explain d-subscript, maybe even better to add stuff like 1-3
days back in? Needs to be comprehensible without recurring to appendix} We differentiate
between an initial period of infection without being infectious or showing symptoms,
being infectious (presymptomatic or asymptomatic), showing symptoms, requiring intensive
care, and recovery or death \citep[similar to][]{Grimm2021}.\comment[id=HM]{Would the
model in \citet{Aleta2020} work?} The probabilities of transitioning between these
states depend on age; their duration is random and calibrated to medical literature (for
a detailed description see Supplementary Material~\ref{sub:data_course_of_disease}). Conditional on the
type of contact, infectiousness is independent of age
\citep{Jones2021}.\comment[id=K]{would be $\beta_{c}$}

The model includes several other features, which are crucial to describe the evolution
of the pandemic in 2020-2021. New virus strains with different infectiousness
profiles\comment[id=K]{$\sigma_v$} may appear. Vaccines may become available. During the
vaccine roll-out, priority may depend on age and occupation; vaccine hesitancy is
modelled by some individuals refusing vaccination offers. With some probability,
vaccinated agents become immune and do not transmit the virus \citep{Hunter2021,
    LevineTiefenbrun2021, Petter2021, Pritchard2021}.

We include two types of tests. Polymerase chain reaction (PCR) tests reveal whether an
individual is infected or not; there is no uncertainty to the result. PCR tests require
at least one day to be processed\comment[id=K]{$\gamma_{PCR,\:d}$ where $d=$ number of
days} and there are aggregate capacity constraints. In contrast, rapid antigen tests
yield immediate results. Specificity\comment[id=K]{$p_{negative|not\:infected}$} and
sensitivity\comment[id=K]{"$p_{positive|infected,\;i,\:t}$", where $i$ is the individual
and $t$ is the date to determine time until/since infectiousness} of these tests is set
according to data analyzed in \cite{Bruemmer2021, Smith2021}; sensitivity depends on the
timing of the test relative to the onset of infectiousness. After a phase-in period, all
tests that are demanded will be performed. Figure~\ref{fig:model_rapid_tests} shows our
model for rapid test demand. Schools may require staff and students to be tested
regularly.\comment[id=K]{"$\pi_{teacher,\:t}$" etc.} Rapid tests may be offered by
employers to on-site workers. Individuals may demand tests for private reasons, which
include having plans to meet other people, showing symptoms of CoViD-19, and a household
member having tested positively for the virus. We endow each agent with an individual
compliance parameter. This parameter determines whether she takes up rapid
tests.\footnote{Positive test results or symptoms leads most individuals to reduce their
    contacts; this is why tests impact the actual contacts in
    Figure~\ref{fig:model_description}.}

Modelling a population of agents according to actual demographic characteristics means
that we can use a wide array of data to identify and estimate the model's many
parameters.\footnote{See Supplementary Material~\ref{sec:data_and_parameters} for a
    complete description.} Contact diaries yield pre-pandemic distributions of contacts for
different contact types and their assortativity by age group. Mobility data is used to
model the evolution of work contacts. School and daycare policies can be incorporated
directly from official directives. Administrative records on the number of tests,
vaccinations by age and region, and the prevalence of virus strains are generally
available. Surveys may ask about test offers, propensities to take them up, and past
tests. Other studies' estimates of the seasonality of infections can be incorporated
directly. The remaining parameters---most notably, these include infection probabilities
by contact network and the effects of some NPIs, see Supplementary
Material~\ref{subsec:estimated_params}---will be chosen numerically so that the model
matches features of the data \citep[see][for the general method]{McFadden1989}. In our
application, we keep the number of free parameters low in order to avoid overfitting.
The data features to be matched include official case numbers for each age group and
region, deaths, and the share of the B.1.1.7 strain.

The main issue with official case numbers is that they will contain only a fraction of
all infections. In the German case, this specifically amounts to positive PCR tests. We
thus model recorded cases as depicted in Figure~\ref{fig:model_official_cases}. We take
mortality-based\comment[id=HM]{added ``mortality-based'' -- correct?} aggregate
estimates of the share of detected cases \comment[id=K]{$\psi$} and use data on the
share of PCR tests administered because of CoViD-19
symptoms\comment[id=K]{$\chi_{symptom}$}. As the share of asymptomatic individuals
varies by age group, this gives us age-specific shares (see
Figure~\ref{fig:share_known_cases_by_age_group} for the share of detected cases by age
group over time in our model). Our estimates suggest that---in the absence of rapid
testing---the detection rate is 80\% higher on average for individuals above age 80
compared to school age children. Once rapid test become available, confirmation of a
positive result\comment[id=K]{prob of demanding confirmation is $\chi_{confirmation}$}
is another reason leading to positive PCR tests.

The model is applied to the second and third waves of the CoViD-19 pandemic in Germany,
covering the period mid-September 2020 to the end of May 2021.
Figure~\ref{fig:pandemic_drivers_model_fit} describes the evolution of the pandemic and
of its drivers. The black line in Figure~\ref{fig:aggregated_fit} shows officially
recorded cases; the black line in Figure~\ref{fig:stringency_infectious_contacts} the
Oxford Response Stringency Index \citep{Hale2020}, which tracks the tightness of
non-pharmaceutical interventions. The index is shown for illustration of the NPIs, we
never use it directly. For legibility reasons, we transform the index so that lower
values represent higher levels of restrictions. A value of zero means all measures
incorporated in the index are turned on. The value one represents the situation in
mid-September, with restrictions on gatherings and public events, masking requirements,
but open schools and workplaces. In the seven weeks between mid September and early
November, cases increased by a factor of ten. Restrictions were somewhat tightened in
mid-October and again in early November. New infections remained constant throughout
November before rising again in December, prompting the most stringent lockdown to this
date. Schools and daycare centers were closed, so were customer-facing businesses except
for grocery and drug stores. From the peak of the second wave just before Christmas
until the trough in mid-February, newly detected cases decreased by almost three
quarters. The third wave in the spring of 2021 is associated with the B.1.1.7 (Alpha)
strain, which became dominant in March (Figure~\ref{fig:share_b117}).\footnote{B.1.617.2
    (Delta) reached Germany in April but still accounted for less than 5\% of cases at the
    end of our simulation period.} In early March, some NPIs were relaxed; e.g.,
hairdressers and home improvement stores were allowed to open again to the public. There
were many changes in details of regulations afterwards, but they did not change the
overall stringency index.

@@

By March 2021, the set of policy instruments had become much more diverse. Around the
turn of the year, the first people were vaccinated with a focus on older age groups and
medical staff (Figure~\ref{fig:antigen_tests_vaccinations}). Until the end of May, 43\%
had received at least one dose of a vaccine. In late 2020, rapid tests started to
replace regular PCR tests for staff in many medical and nursing facilities. These had to
be administered by medical doctors or in pharmacies. At-home tests approved by
authorities became available in mid-March. Rapid test centers were opened, and one test
per person and week was made available free of charge. In several states, customers were
only allowed to enter certain stores with a recent negative rapid test result. These
developments are characteristic of many countries: The initial focus on NPIs to slow the
spread of the disease has been accompanied by vaccines and a growing acceptance and use
of rapid tests. At broadly similar points in time, novel strains of the virus have
started to pose additional challenges.

We draw simulated samples of agents from the population structure in September 2020 and
use the model to predict recorded infection rates until the end of May 2021. See
Supplementary Materials~\ref{subsec:synthetic_population} and
\ref{sub:initial_conditions} for details. The blue line in
Figure~\ref{fig:aggregated_fit} shows that our model's predictions are very close to
officially recorded cases in the aggregate. This is also true for infections by age and
geographical region (see Supplementary Material~\ref{sec:subsec:fit_results}).

The effects of various mechanisms can be disentangled due to the distinct temporal
variation in the drivers of the pandemic. Next to the stringency index, the three lines
in Figure~\ref{fig:stringency_infectious_contacts} summarize how contact reductions,
increased hygiene regulations, and seasonality evolved since early September for each of
the three broad contact networks. For example, a value of 0.75 for the work multiplier
means that if the environment was the same as in September (levels of infection rates,
no rapid tests or vaccinations, only the wildtype virus present), infections at the
workplace would be reduced by 25\%. Two aspects are particularly interesting. First, all
lines broadly follow the stringency index and they would do so even more if we left out
seasonality and school vacations (roughly the last two weeks of October, two weeks each
around Christmas and Easter, and some days in late May). Second, the most stringent
regulations coincide with the period of decreasing infection rates between late December
2020 and mid-February 2021. The subsequent reversal of the trend is associated with the
spread of the B.1.1.7 variant. During the steep drop in recorded cases during May 2021,
for 42\% of the population took at least one rapid tests per week, the first-dose vaccination
rate rose from 28\% to 43\%, and seasonality lowered the relative infectiousness
of contacts.

In order to better understand the contributions of rapid tests, vaccinations, and
seasonality on the evolution of infections in 2021,
Figure~\ref{fig:2021_scenarios_broad} considers various scenarios. NPIs are always held
constant at their values in the baseline scenario.
Figure~\ref{fig:2021_scenarios_recorded} shows the model fit (the blue line, same as in
Figure~\ref{fig:aggregated_fit}), a scenario without any of the three factors (red
line), and three scenarios turning each of these factors on individually.
Figure~\ref{fig:2021_scenarios_newly_infected} does the same for total infections in the
model. Figure~\ref{fig:2021_scenarios_decomposition} employs Shapley values
\citep{Shapley2016} to decompose the difference in total infections between the scenario
without any of the three factors and our main specification.

@@

Until mid-March, there is no visible difference between the different scenarios.
Seasonality hardly changes, and only few vaccinations and rapid tests were administered.
Even thereafter, the effect of the vaccination campaign is surprisingly small at first
sight. Whether considering recorded or total infections with only one channel active,
the final level is always the highest in case of the vaccination campaign (orange
lines). The Shapley value decomposition shows that vaccinations contribute 16\% to the
cumulative difference between scenarios. Reasons for the low share are the slow
start---it took until March~24th until 10\% of the population had received their first
vaccination, the 20\% mark was reached on April 19th---and the focus on older
individuals. These groups contribute less to the spread of the disease than others due
to a lower number of contacts. By the end of our study period, when first-dose
vaccination rates reached 43\% of the population, the numbers of new cases would have
started to decline. It is important to note that the initial focus of the campaign was
to prevent deaths and severe disease. Indeed, the case fatality was rate considerably
lower during the third wave when compared to the second (4.4\% between October and
February and 1.4\% between March and the end of May).\comment[id=K]{This needs a
source.}\comment[id=HM]{Didn't we calculate that ourselves at some point?}

@@

A similar-sized effect---42\% in the decomposition---comes from rapid testing. Here, it
is crucial to differentiate between recorded cases and actual cases. Additional testing
means that additional infections will be recorded which would otherwise remain
undetected. Figure~\ref{fig:2021_scenarios_recorded} shows that this effect is large and
may persist for some time. Until late April, recorded cases are higher in the scenario
with rapid testing alone when compared to the setting where none of the three mechanisms
are turned on. The effect on total cases, however, is visible immediately in
Figure~\ref{fig:2021_scenarios_newly_infected}. Despite the fact that only 10\% of the
population performed weekly rapid tests in March on average, new infections on April~1
would have been reduced by 53\% relative to the scenario without vaccinations, rapid
tests, or seasonality.

So why is rapid testing so effective? In order to shed more light on this question,
Figure~\ref{fig:2021_scenarios_decomposition_tests} decomposes the difference in the
scenario without rapid tests only and the main specification into the three channels for
rapid tests. Tests at schools have the smallest effect, which is largely explained by
schools not operating at full capacity during our period of study and the relatively
small number of students.\footnote{18\% of our population are in the education sector
    (pupils, teachers, etc.); 46\% are workers outside the education sector.} Almost 40\%
come from tests at the workplace. Despite the fact that rapid tests for private reasons
are phased in only in mid-March, they make up for more than half of the total effect.
The reason lies in the fact that a substantial share of these tests is driven by an
elevated probability to carry the virus, i.e., showing symptoms of CoViD-19 or following
up on a positive test of a household member. The latter is essentially a form of contact
tracing, which has been shown to be very effective \citep{Contreras2021,
    Fetzer2021,Kretzschmar2020}. Indeed, a deeper analysis in Supplementary
Material~\ref{subsec:appendix_scenarios} shows that the same amount of rapid tests
administered randomly in the population would not nearly have been as effective.

@@

Our analysis has shown that during the transition to high levels of vaccination and
possibly thereafter, large-scale rapid testing can substitute for some NPIs. This comes
at a fraction of the cost. A week of the fairly strict lockdown in early 2021 is
estimated to have cost around 20~Euros per capita \citep{Wollmershauser2021}; retail
prices for rapid tests were below one Euro in early June 2021. While we do not
distinguish between self-administered rapid tests and point of care rapid tests, the
former are likely to play a larger role for indication-driven testing. Widespread
availability at low prices seems important. However, they rely on purely voluntary
participation in a non-public setting. The benefit of point-of-care rapid tests as a
precondition to participate in leisure activities as well as mandatory tests at the
workplace or at school come from screening the entire population. This is important
because disadvantaged groups are less likely to be reached by testing campaigns relying
on voluntary participation \citep[e.g.][]{StillmanTonin2021}; at the same time, these
groups have a higher risk to contract CoViD-19 \citep{KochInstitut2021a}. Mandatory
tests at school and at the workplace will extend more into these groups. The same goes
for individuals who exhibit a low level of compliance with CoViD-19-related regulations.
Compared to vaccinations, rapid testing programmes allow a much quicker roll-out, making
it arguably the most effective tool to contain the pandemic in the short run.
